---
title: "Hotspot identification using a relative frequency distribution approach"
author: "PJ Bouchet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	message = TRUE,
	warning = TRUE,
	collapse = TRUE,
	comment = "#>"
)
```

This small vignette illustrates the use of custom functions for identifying 'hotspots' in a spatial variable of interest. 

The code is based on the frequency distribution approach proposed by Bartolino et al. (2011)^[Bartolino V, Maiorano L, Colloca F. 2011. A frequency distribution approach to hotspot identification. Population Ecology, 53(2): 351-359] and applied in Bouchet et al. (2017).^[Bouchet et al. 2017. Continental-scale hotspots of pelagic fish abundance inferred from commercial catch records. Global Ecology and Biogeography, 26(10): 1098-1111.]

The required libraries and options are specified below.

```{r message=FALSE, warning=FALSE, tidy=TRUE}
library(fANCOVA)
library(raster)
library(virtualspecies)
library(tidyverse)

options(tibble.width = Inf) # All tibble columns shown
options(pillar.neg = FALSE) # No colouring negative numbers
options(pillar.subtle = TRUE)
options(pillar.sigfig = 4)
```

Below is a simple simulated example using WorldClim data.
WorldClim (www.worldclim.org) freely provides gridded climate data (temperature and precipitations) for the entire World. Here, we generate a virtual tropical species (ie. living in hot and humid environments), whose response to two environmental variables, the annual mean temperature bio1 and annual mean precipitation bio2, is pre-determined. 
Code adapted from: http://borisleroy.com/files/virtualspecies-tutorial.html#input-data 

```{r tidy=TRUE, fig.width=7, fig.height=5}
# Downloads WorldClim data
worldclim <- getData("worldclim", var = "bio", res = 10)

# Gaussian functions of bio1 and bio12
my.parameters <- formatFunctions(bio1 = c(fun = 'dnorm', mean = 250, sd = 50),
                                 bio12 = c(fun = 'dnorm', mean = 4000, sd = 2000))

# Generates simulated species distributions
simsp <- generateSpFromFun(raster.stack = worldclim[[c("bio1", "bio12")]],
                                      parameters = my.parameters,
                                      plot = TRUE)

# For the sake of the example only extracts a small portion of that distribution
# eg from South America
simdist <- raster::crop(x = simsp$suitab.raster,
                        y = as(extent(c(-51,-45, -15,-5)), "SpatialPolygons"))

plot(simdist)
# Turns the raster into a data.frame object and manipulates values
dat <- raster::as.data.frame(simdist, xy=T)
dat <- dat %>% purrr::set_names(., c("longitude", "latitude", "z"))
dat$z = round(dat$z *100,0)

```

Next we define custom functions to calculate the cumulative relative frequency distribution function CRFD (as described in the Bartolino paper)

```{r echo=TRUE, tidy=TRUE}
no.cumul<-function(data, value){
  return(length(data[data<value]))}

bartolino<-function(df, var.name){
  
  # Converts var.name to a string
  var.name <- deparse(substitute(var.name))
  
  # Calculates x-axis [x = z / zm] and y-axis [y = M(x)/n] values 
  df <- df %>% 
    mutate_(x = paste(var.name, "/max(", var.name, ")"))
  
  allx <- df %>% pull(x)
    
  df <- df %>% rowwise %>%
    mutate(y=no.cumul(allx, x)/nrow(df)) %>%
    ungroup()

  return(df)} # End function
```

These functions can be applied to the data.

```{r echo=TRUE, tidy=TRUE, warning=FALSE}
dat.bart <- bartolino(df = dat, var.name = z)
```

Additional functions are needed to determine the highest x-value corresponding to a 45 degree tangent to the predicted loess curve.

```{r tidy=TRUE}
find.bartolino<-function(loess.predictions){
  
  for (p in 3:(length(loess.predictions)-1)){
    slope.curve<-(loess.predictions[p+1]-loess.predictions[p-1])/(loess.x[p+1]-loess.x[p-1])
    if(slope.curve>=1) {
      barto.x<-loess.x[p]
      break
    }}
  return(barto.x)
} # End function

find.bartolino.y<-function(loess.predictions){
  
  for (p in 3:(length(loess.predictions)-1)){
    slope.curve<-(loess.predictions[p+1]-loess.predictions[p-1])/(loess.x[p+1]-loess.x[p-1])
    if(slope.curve>=1) {
      barto.y<-loess.predictions[p]
      break
    }}
  return(barto.y)
} # End function
```

A LOESS model can be fitted to the CRFD points, and the x-value cutoff identified from the resulting predictions.

```{r tidy=TRUE, warning=FALSE, fig.width=7, fig.height=5}
# LOESS smoothing - with automated span selection based on AICc or GCV
# The degree of smoothing can also be selected manually using user.span
  
  loess.x<-rev(seq(0,1,0.001))
  loessmod.n<- fANCOVA::loess.as(dat.bart$x, 
                        dat.bart$y, 
                        degree = 0, 
                        criterion = c("aicc", "gcv")[1],
                        user.span = NULL, 
                        plot = T)
  
  lo.preds.n<-predict(loessmod.n,loess.x)
  
  xbest<-find.bartolino(lo.preds.n)
  ybest<-find.bartolino.y(lo.preds.n)
  abline(v=xbest, col="orange", lty=2)
```

Finally, hotspots are simply those sites with an x value higher than the threshold.

```{r warning=FALSE, fig.width=7, fig.height=5, tidy=TRUE}
dat.hot <- dat.bart %>% 
  filter(x>=xbest)

# Converts to SpatialPoints df
hot.pts <- dat.hot %>% 
  dplyr::select(longitude, latitude, z) %>% 
  SpatialPointsDataFrame(coords = cbind(.$longitude, .$latitude), data = .)

# Visualisation
plot(simdist)
plot(hot.pts, add=T)

```

